Relationships in the extremes and their influence on competition and coexistence: Supporting information

### 2022-01-26 paper/SupMat.Rnw
edits... stopping for now

%***DAN: Jasmin, please start on this section when you get a chance. Here are some topics that need
%to be covered:
Estimations of $Q_i(t)$, $V_i(t)$, and $K_i(t)$, temperature dependence functions for 
cell quota, reproduction rate, and half saturation, respectively, were used to 
paramaterize the competing diatom model outline in the paper. 
Values of and subsequent function estimations are plotted in Figure \ref{fig:VQK}. 
\cite{Descamps_2005} obtained estimated values for $Q_i$, $V_i$, and $K_i$ from 
single species nine-day batch experiments at four different temperatures ($6^\circ$C, 
$12^\circ$C, $18^\circ$C, $24^\circ$C) (see Table 1 of \cite{Descamps_2005}) and further
computed a linear interpolation to define temperature dependence 
functions of $Q_i(t)$, $V_i(t)$, and $K_i(t)$. These values and functions are 
indicated in Figure \ref{fig:VKQ} by open circles and dashed lines. 
\cite{Ellne_2016, Ellner_2019} chose to slightly modify some values and functions 
(to better align model with competition experiments), 
%why did they make K_i's constant and do we need to adjust so that it
%is constant over the range of temperatures considered 
indicated in Figure \ref{fig:VKQ} by closed circles and solid lines. 
These modifications do not result in meaningful differences when evaluating 
coexistence at the parameter values used by \cite{Descamps_2005}.
We chose to implement the modified functiones in our methods to quantify and 
decompose coexistence.

@@

Our decomposition is defined by $E(t)=\theta(t)$, 
$C_j(t)=(K_j(t)+S(t))/S(t)$ and $r_j(E_j,C_j)=\frac{V_j(E)}{C_j} -D$. 
We note that our definition and selection of our "features" of interest are
options among many and features need not be limited to $E$ and $C$ \citep{Ellner_2019}. 
For example \cite{Ellner_2019} decomposes coexistence with respect to fluctuation in
inflow silicate concentration, $S$, and temperature, and \cite{Ellner_2016} decomposes
coexistence using classical $E$ and $C$ drivers, but define $E=V(t)$. 
Choice of drivers and their definitions are up to the investigator,
but should real variables that influence population growth.
Thus in our exploration, environmental fluctuation is temperature fluctuation.

%3) You need to outline how the quantities E_i and C_i\i were obtained from simulations.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combine to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.

>>>>

%***DAN: Jasmin, please start on this section when you get a chance. Here are some topics that need
%to be covered:
Estimations of $Q_i(t)$, $V_i(t)$, and $K_i(t)$, temperature dependence functions for 
cell quota, reproduction rate, and half saturation, respectively, were used to 
paramaterize the competing diatom model outline in the paper. 
Values of and subsequent function estimations are plotted in Figure \ref{fig:VQK}. 
\cite{Descamps_2005} obtained estimated values for $Q_i$, $V_i$, and $K_i$ from 
single species nine-day batch experiments at four different temperatures ($6^\circ$C, 
$12^\circ$C, $18^\circ$C, $24^\circ$C) (see Table 1 of \cite{Descamps_2005}) and further
computed linear interpolations to
functions $Q_i(t)$, $V_i(t)$, and $K_i(t)$. These values and functions are 
indicated in Figure \ref{fig:VKQ} by open circles and dashed lines. 
\cite{Ellne_2016, Ellner_2019} chose to slightly modify some values and functions 
(to better align model with competition experiments), 
%why did they make K_i's constant and do we need to adjust so that it
%is constant over the range of temperatures considered 
indicated in Figure \ref{fig:VKQ} by closed circles and solid lines. 
These modifications do not result in meaningful differences when evaluating 
coexistence at the parameter values used by \cite{Descamps_2005}.
We chose to implement the modified functiones in our methods to quantify and 
decompose coexistence.

@@

Our decomposition is defined by $E(t)=\theta(t)$, 
$C_i(t)=(K_i(t)+S(t))/S(t)$ and $r_i(E_i,C_i)=\frac{V_i(E)}{C_i} -D$. 
We note that our definition and selection of our "features" of interest are
options among many and features need not be limited to $E$ and $C$ \citep{Ellner_2019}. 
For example \cite{Ellner_2019} decomposes coexistence with respect to fluctuation in
inflow silicate concentration, $S$, and temperature, and \cite{Ellner_2016} decomposes
coexistence using classical $E$ and $C$ drivers, but define $E=V(t)$. 
Like \cite{Ellner_2016}, we set $C_i(t)=(K_i(t)+S(t))/S(t)$ to reflect 
potential versus actual growth of the diatom populations. 
Choice of drivers and their definitions are up to the investigator,
but should refelct real variables that influence population growth.

%3) You need to outline how the quantities E_i and C_i\i were obtained from simulations. 
To obtain $E_i$ and $C_{i \bs i}$, the model was simulated for a $T=3600$ number of timesteps, setting initial population size of invader i to be zero and initial population of the resident, $r=20$.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.
